\documentclass{llncs}

\title{Combining Modeling and Programming\thanks{The work
  described in this publication was carried out at Jet Propulsion
  Laboratory, California Institute of Technology, under a contract with
  the National Aeronautics and Space Administration.}}
\author{Klaus Havelund \and Rahul Kumar}
\institute{
  Jet Propulsion Laboratory\\
  California Institute of Technology\\
  California, USA
}


\input{macros}

\begin{document}

\maketitle

\input{abstract}

%%% TABLE OF CONTENTS - UNDER DEVELOPMENT: %%%

\vspace{1cm}
\color{red}
\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}\\

{\bf Table of Contents}

\begin{enumerate}
  \item Introduction: Discuss what 'change' means, and how it is the
    same idea as 'refinement'. Discuss 'modeling' and 'programming',
    and how they are becoming one. Our contributions. 
  \item The K Language: Present the K language. Discuss motivation,
    influences, syntax, ability and provide an example. Integration
    with Graphical Modeling (???). Verification with K.
  \item Change. What is it? How does one deal with it? What is our
    idea of dealing with change? Start presenting examples and
    discussing how they can be verified using K. For each example,
    show the \emph{old} and \emph{new} version of the program, discuss
    the differences, discuss what K does and what can be inferred from
    that.
  \item Related Work
  \item Conclusion
\end{enumerate}

\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}\\
\color{black}
\vspace{1cm}

%%%%%% 

\input 	{introduction}

\input{k}

\input{change}

\input{conclusion}

\bibliographystyle{abbrv}
\bibliography{biblio}

\end{document}
